---
title: "HW 4"
output: html_document
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(car, here, tidyverse, dplyr, ggplot2, data.table, lubridate, glmnet, stargazer, forecast, lmtest, hwwntest)
```

## R Markdown

The file ChinaImports2.txt gives monthly imports by the U.S. from China, in millions of 
dollars, for the period January 1989 through December 2019.

1.  Graph both the monthly imports and the logged monthly imports vs. time.  Comment 
in detail on what you can determine from the plots.  Mark economic downturns in the 
plots.

**Discussion:** 

**TREND** - Upward trend is visible. In the imports plot (blue line), imports seem to increase exponentially or by a constant multiplicative factor until the 2009 recession period (highlighted in orange) where it begins to flatten. In the log Imports plot (red line) log Imports seem to increase by a constant additive amount before flattening after the 2009 recession. 

**SEASONALITY** - The plots indicate strong seasonality in imports. Imports tend to reach its highest peak around October annually. After dropping in November and December, Imports tend to increase in January, making a small peak before dropping again in February. From February to October, imports tend to increase at a steady rate.

**VOLATILITY** - There is evidence of increasing volatility in the imports plot. For example, it is evident that imports tend to rise and drop more sharply after the 2009 recession period. This suggest a multiplicative model is preferred than an additive one.

**IMPACT OF ECONOMIC DOWNTURN** - Economic downturn tend to cause imports to drop, as is evident in the 2001 and 2009 contractions (the drop is less prominent in 1990 where imports from China is still very low compared to the following periods).

**OUTLIER** - The drop in imports in February of 2009 is likely an outlier.

```{r}
f <- file.choose("ChinaImports2.txt")
imports <- read.csv(f)
```
```{r}
imports <- imports %>% mutate (year_month = paste(Year,Month,sep="-"))
```

```{r}
# modified
cycle <- data.frame(from = c('1945-2','1948-11','1953-7','1957-8','1960-4','1969-12','1973-11','1980-1','1981-7','1990-7','2001-3','2007-12','2020-3'),
                    to = c('1945-10','1949-10','1954-5','1958-4','1961-2','1970-11','1975-3','1980-7','1982-11','1991-3','2001-11','2009-6','2020-4'))
```

```{r}
#cycle <- data.frame(from = #c('1945-02-01','1948-11-01','1953-07-01','1957-08-01','1960-04-01','1969-12-01','1973-11-01','1#980-01-01','1981-07-01','1990-07-01','2001-03-01','2007-12-01','2020-03-01'),
#                    to = #c('1945-10-01','1949-10-01','1954-05-01','1958-04-01','1961-02-01','1970-11-01','1975-03-01','1#980-07-01','1982-11-01','1991-03-01','2001-11-01','2009-06-01','2020-04-01'))
```

```{r echo=FALSE}
# we create a new column "rects" which signals 1 if it row corresponds to a "recession" period according to NEBR data

lst <- rep(0, nrow(imports))
imports$rects <-lst

imports<-transform(imports,rects=ifelse(year_month %in% cycle$from,1,rects))
imports<-transform(imports,rects=ifelse(year_month %in% cycle$to,2,rects))

for (i in seq(2, nrow(imports), by=1)) {
  if (imports$rects[i] == 0) {
    if (imports$rects[i-1] == 1) {
       imports$rects[i] <- 1
    }
  }
}

imports<-transform(imports,rects=ifelse(rects==2,1,rects))
```

```{r echo=FALSE}
imports <- imports %>% mutate(Time = 1:n())
```

```{r echo=FALSE}
set.seed(0)
dat <- imports

## Determine highlighted regions
v <- dat$rects

## Get the start and end points for highlighted regions
inds <- diff(c(0, v))
#print(v)
start <- dat$Time[inds == 1]
end <- dat$Time[inds == -1]
if (length(start) > length(end)) end <- c(end, tail(dat$Time, 1))

## highlight region data
rects <- data.frame(start=start, end=end, group=seq_along(start))

ggplot(data=dat, aes(Time, Imports)) +
theme_minimal() +
geom_line(color = "#00AFBB", size = .5) +
geom_rect(data=rects, inherit.aes=FALSE, aes(xmin=start, xmax=end, ymin=min(dat$Imports),
ymax=max(dat$Imports), group=group), color="transparent", fill="orange", 
alpha=0.3) +
labs(title = "Monthly U.S. Imports from China (in millions of 
dollars)", subtitle = "(From 1989 to 2019 - Contraction period highlighted in Orange)") 
```
```{r}
ggplot(data=dat, aes(Time, logImports)) +
theme_minimal() +
geom_line(color = "#FC4E07", size = .5) +
geom_rect(data=rects, inherit.aes=FALSE, aes(xmin=start, xmax=end, ymin=min(dat$logImports),
ymax=max(dat$logImports), group=group), color="transparent", fill="orange", 
alpha=0.3) +
labs(title = "Monthly U.S. LOG Imports from China (in millions of 
dollars)", subtitle = "(From 1989 to 2019 - Contraction period highlighted in Orange)") 
```
2.  Fit a regression model with seasonal and significant calendar components to the 
differences of the logged monthly imports.  
``` {r}
# add dummy variable for outlier
feb09<-rep(0, max(imports$Time))
feb09[242] <- 1
imports<-data.frame(imports,feb09)
```

```{r}
# add fMonth variables
imports <- imports %>% mutate(fMonth = as.factor(Month))
```

```{r}
# add differences column
logImports_lst <- imports$logImports
differences <- diff(logImports_lst,lag=1)
imports$differences <- differences

```
```{r}
model1<-lm(logImports~Time+I(Time^2)+I(Time^3)+fMonth+feb09+c348+s348+c432+s432, data = imports);
summary(model1)
```

(a)  Describe your fitted model.

(b)  Tabulate and plot the estimated seasonal indices for the imports series.  Provide 
careful interpretation of the estimates.
(c)  Perform a residual analysis for your model, examining the plot of the residuals vs. 
time, a residual normal quantile plot, the residual acf and pacf plots, and the residual 
spectral density (along with Bartlett’s test).  Discuss all.
3.  (a)  Fit a regression model with just the calendar variables to the logged monthly 
imports series.  [The calendar pairs will not be significant in this model.  That’s okay—
do so and proceed as below.]
(b)  Fit a seasonal ARIMA model to the residuals from the model in part (a). 
(i)  Describe your fitted ARIMA model.
(ii)  Tabulate and plot estimated static seasonal indices.  Compare the static estimates to 
those obtained via regression in part 2(b), with a table and a plot.  Present the dynamic 
seasonal index estimates graphically as with the examples in the notes.  Discuss the 
dynamic estimates.  What information do they provide about changes over time of 
seasonal structure?
(iii)  Perform a residual analysis for your ARIMA model, examining the plot of the 
residuals vs. time, the residual normal quantile plot, the residual acf and pacf, and the 
residual spectral density (along with Bartlett’s test).  Has the ARIMA model you’ve fit 
produced reduction to white noise?  Discuss carefully.
4.  Give some brief concluding remarks.  What has the analysis revealed about imports 
from China to the U.S. during the years 1989 to 2019?

